---
title: "The final project - Animated changes of GDP PPP 1990-2018"
author: "Bartosz Czernecki"
date: "1/14/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overall aim
The animated charts are increasingly interested in many social media podcasts. 
For example, the popularity among programming languages is shown in this link: <https://www.youtube.com/watch?v=Og847HVwRSI>

In the final project we will try to create an animated chart that will show us changes of 
Growth Domestic Product (GDP) standardized by Purchasing Power Paritity (PPP) in all countries that participate in our course. The values are given in USD.

# Dataset

The World Bank publishes GDP PPP reports on annual basis  
(<https://data.worldbank.org/indicator/NY.GDP.PCAP.PP.CD>) for all countries around the world. This dataset has been pre-formated to the form readable in spreadsheet-like file.

Please download the file from <http://iqdata.pl/dataprocessing/data/> and save it in a known location on your laptop. 

## Reading dataset

Set your working directory to the location of downloaded file. Please use from the upper
menu `SESSION` then `SET WORKING DIRECTORY` and `CHOOSE DIRECTORY`.

Activate library that allows reading "XLSX" format (e.g. `readxl`) and name the read data as `df` object. The structure of the created `df` object should be as shown below for the first 6 rows:
```{r reading, echo=FALSE}
library(readxl)
df = read_excel("data/gdp_ppp.xlsx")
head(df)
```

## Re-shaping data 

In the final phase of the project we will use `ggplot2` package to create our charts. This
package requires so called narrow format of the data frame. Therefore our data frame needs
to be reshaped to the form where we will have 4-5 columns with: `country`, `code`, `year` and `gdp`.

It can be done *manually* or with the use of a pivot-like functions. 
One of my favourie is the `gather` function from the `tidyr` package, which probably will have to be installed on your laptop if not used before. 

Activate the `tidyr` and `dplyr` (or `tidyverse`) libraries and launch the `?gather` command to familiarize yourself with the way it works. Try to reshape our `df` object to the following structure:


```{r reshape, echo=FALSE}
library(dplyr)
library(tidyr)
df2 = gather(df, key = year,  value = "gdp", 3:31)
```

```{r reshape2, echo=TRUE}
head(df2)
```

Hints: Use the `year` column as the **key** argument, and `value = "gdp"`. Save the obtained results as `df2`.

## Choosing countries

Right now our data frame `df2` contains data for all countries. We'd like to limit our
animation only to countries that participate in our class, i.e.: Czech Rep., Germany, Poland South Korea and Spain. Additionally, we'd like to add average for the entire world which is given as *WLD* in the column `code`.

Find out codes for the chosen countries and use the `filter` command to clip the dataset for these 6 codes. Fill in the following command and save results as `gdp_tidy`

```{r filter}
# gdp_tidy = filter(df2, code %in% c("COUNTRY_1", "COUNTRY_2", ...) )
```

This is the first 10 rows of our `gdp_tidy` should return:
```{r filter1, echo=FALSE}
gdp_tidy = filter(df2, code %in% c("POL", "DEU", "CZE", "KOR", "ESP", "WLD"))
head(gdp_tidy, 10)
```